### 2024-09-18 Paper/MSEPM-Manuscript.tex
Update MSEPM-Manuscript.tex
more minor edits

The site clusters largely conform to underlying biological expectations. Cluster one contains sites 
that are wholly associated with sex status and localized to the X chromosome (Supp. Table 1) and is enriched for peaks of 
transcription factors associated with sex specific regulation such as MAZ\cite{Lopes-Ramos2020-ex}. 
Clusters nine and ten contain sites whose methylation status is largely driven by CT-PC1, and are enriched for 
transcription factors associated with granulocyte development (CEBPB, CEBPA, EP300, ETV6)\cite{Theilgaard-Monch2022-zw,Guerzoni2006-ii}. 
Similarly, clusters two, five and eight are associated with CT-PC2 and are enriched for 
transcription factor peaks associated with immune development (ZBED1, ETV6, FOSL2, FOS, TBX21). Clusters four and 
six are associated with loss of methylation with age. Cluster six is highly enriched for CTCF binding sites; CTCF 
is known to increase at sites where methylation is lost during aging\cite{Tharakan2020-pj}. Cluster four is enriched 
for STAT3 whose activation during exercise is age dependent\cite{Mohamed2020-he,Trenerry2008-kj}.  
Cluster seven is associated with the accumulation of methylation with age and is enriched for immunoprecipitation peaks
 for aging associated transcription factors SMAD4 and RE1-Silencing Transcription Factor (REST). SMAD4 encodes a protein 
 involved in the transforming growth factor beta (TGF-$\beta$) signaling pathway. Age related dysregulation of TGF-$\beta$ 
 has been linked to reduced skeletal muscle regeneration\cite{Carlson2009-uz,Paris2016-fo} and SMAD4 polymorphisms 
 are associated with longevity\cite{Carrieri2004-by}). REST is a transcriptional repressor of neuron specific genes in 
 non-neuronal cells\cite{Chong1995-dj,Coulson2005-pb}. REST expression is upregulated in aged prefrontal cortex 
 tissue and the absence of REST expression is associated with cognitive impairment\cite{Lu2014-dz} and 
 cellular senescence in neurons\cite{Rocchi2021-od}).

@@

Based on these results we hypothesize that the mechanisms that underlies the gain of methylation with age at these 
bivalent promoters is the age-associated loss of H3K4me3.  It is well established that the presence of trimethylation
 on H3K4 inhibits de novo methylation, and this effect explains the hypomethylation that is typical of promoters, 
 including bivalent promoters.  We therefore hypothesize that the gain of methylation at these sites may be caused 
 by an age associated loss of H3K4me3.  In order to demonstrate that H3K4me3 decreases with age for genomic regions 
 where DNA methylation increases, we used published H3K4me3 ChIP-seq data from epigenomesportal\cite{Bujold2016-vk}. 
 We selected two different blood cell types CD38+ B Cells and CD56+ NTK Cells and plotted the H3K4me3 signal of young 
 (0 to 5 years old) versus old individuals (60 to 65 years old) over genomic regions of cluster 7 (Figure 7A). 
 Our analysis shows that younger individuals have higher levels of H3K4me3 compared to older ones (Figure 7A) as 
 also shown for two selected genomic loci of cluster 7 (the promoters of KCTD1 and IRS2 genes) where we can observe
 a marked decrease in the levels of H3K4me3 as age increases (Figure 7B). All together these data suggest that genomic 
 regions whose DNA methylation is increased with age exhibit an age dependent loss of H3K4me3, thus showing an inverse 
 correlation between DNA methylation and H3K4me3 at these genomic loci.

@@

In conclusion, we introduced a multi-dimensional extension of the Epigenetic Pacemaker, the MSEPM. The MSEPM is capable of 
accurately modeling multiple methylation associated factors simultaneously. This paradigm can elucidate the site specific 
regulation underpinning methylome dynamics. It allows us to characterize the mechansims underlying age assocated increases 
in methylation sites, suggeting that these were caused by the loss of H3K4Me3 at bivalent promoters of genes that are silenced 
in blood.  The MSEPM is available under the MIT license at https://github.com/NuttyLogic/MultistateEpigeneticPacemaker.

>>>>

The site clusters largely conform to underlying biological expectations. Cluster one contains sites 
that are wholly associated with sex status and localized to the X chromosome (Supp. Table 1) and is enriched for peaks of 
transcription factors associated with sex specific regulation such as MAZ\cite{Lopes-Ramos2020-ex}. 
Clusters nine and ten contain sites whose methylation status is largely driven by CT-PC1, and are enriched for 
transcription factors associated with granulocyte development (CEBPB, CEBPA, EP300, ETV6)\cite{Theilgaard-Monch2022-zw,Guerzoni2006-ii}. 
Similarly, clusters two, five and eight are associated with CT-PC2 and are enriched for 
transcription factor peaks associated with immune development (ZBED1, ETV6, FOSL2, FOS, TBX21). Clusters four and 
six are associated with loss of methylation with age. Cluster six is highly enriched for CTCF binding sites; CTCF 
is known to increase at sites where methylation is lost during aging\cite{Tharakan2020-pj}. Cluster four is enriched 
for STAT3 whose activation during exercise is age dependent\cite{Mohamed2020-he,Trenerry2008-kj}.  
Cluster seven is associated with the accumulation of methylation with age and is enriched for immunoprecipitation peaks
 for aging associated transcription factors SMAD4 and RE1-Silencing Transcription Factor (REST). SMAD4 encodes a protein 
 involved in the transforming growth factor beta (TGF-$\beta$) signaling pathway. Age related dysregulation of TGF-$\beta$ 
 has been linked to reduced skeletal muscle regeneration\cite{Carlson2009-uz,Paris2016-fo} and SMAD4 polymorphisms 
 are associated with longevity\cite{Carrieri2004-by}. REST is a transcriptional repressor of neuron specific genes in 
 non-neuronal cells\cite{Chong1995-dj,Coulson2005-pb}. REST expression is upregulated in aged prefrontal cortex 
 tissue and the absence of REST expression is associated with cognitive impairment\cite{Lu2014-dz} and 
 cellular senescence in neurons\cite{Rocchi2021-od}.

@@

Based on these results we hypothesize that the mechanisms that underlie the gain of methylation with age at these 
bivalent promoters is the age-associated loss of H3K4me3.  It is well established that the presence of trimethylation
 on H3K4 inhibits de novo methylation, and this effect explains the hypomethylation that is typical of promoters, 
 including bivalent promoters.  We therefore hypothesize that the gain of methylation at these sites may be caused 
 by an age associated loss of H3K4me3.  In order to demonstrate that H3K4me3 decreases with age for genomic regions 
 where DNA methylation increases, we used published H3K4me3 ChIP-seq data from epigenomesportal\cite{Bujold2016-vk}. 
 We selected two different blood cell types CD38+ B Cells and CD56+ NTK Cells and plotted the H3K4me3 signal of young 
 (0 to 5 years old) versus old individuals (60 to 65 years old) over genomic regions of cluster 7 (Figure 7A). 
 Our analysis shows that younger individuals have higher levels of H3K4me3 compared to older ones (Figure 7A) as 
 also shown for two selected genomic loci of cluster 7 (the promoters of KCTD1 and IRS2 genes) where we can observe
 a marked decrease in the levels of H3K4me3 as age increases (Figure 7B). All together these data suggest that genomic 
 regions whose DNA methylation is increased with age exhibit an age dependent loss of H3K4me3, thus showing an inverse 
 correlation between DNA methylation and H3K4me3 at these genomic loci.

@@

In conclusion, we introduced a multi-dimensional extension of the Epigenetic Pacemaker, the MSEPM. The MSEPM is capable of 
accurately modeling multiple methylation associated factors simultaneously. This paradigm can elucidate the site specific 
regulation underpinning methylome dynamics. It allows us to characterize the mechansims underlying age assocated increases 
in methylation sites, suggesting that these were caused by the loss of H3K4Me3 at bivalent promoters of genes that are silenced 
in blood.  The MSEPM is available under the MIT license at https://github.com/NuttyLogic/MultistateEpigeneticPacemaker.
